Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-10 draft.tex
Merged in subsection 6.6.2- "UNSCR-1540 Nuclear Security Standards." Sending this commit to Denia.

@@

@@

@@

RPMs evolved out of a layered defense concept to prevent a low-probability/high-impact scenario, nuclear terrorism. This layered defense is composed of RPMs as discussed in \autoref{sec:domestic_protection}, the overseas equivalent known as the SLD as discussed in \autoref{sec:Trafficking}, and the Global Threat Reduction Initiative discussed in

\section{??????? Insert cross-ref here ??????}

6.6. something

Section XXX INSERT REF TO NUCLEAR THREAT REDUCTION INITIATIVE SECTION HERE XXX.  Despite the layered technical and diplomatic approach, there are significant gaps in the ability to intercept a shipment into the U.S. containing SNM.  For example, DHS has implemented 99\% screening at seaports, but struggles to implement a significant screening at airports, due to a lack of natural choke points and disruption caused by screenings, and at the U.S.'s notoriously porous border crossings \cite{DepartmentofHomelandSecurityDHS2013,GovernmentAccountabilityOffice2012}.  Even if a shipment does cross a border, seaport, or airport with a RPM, the detection probability is only 90\% in the best of scenarios, if the RPM is even operational at the time \cite{DepartmentofHomelandSecurityDHS2013,DepartmentofHomelandSecurityDHS2006}.  To top of the list of issues, the ANSI standard for certification of RPMs contains no SNM requirements due to the inherent difficult in passive detection of SNM \cite{Ansi2007}.  The ASP program was conceived to address some of these problems and increase interception through advanced detectors and algorithms, but it failed to pass 6 of 10 criteria in field tests, criteria which were widely criticized as being favorable to the vendors in the first place \cite{UnitedStatesGovernmentAccountabilityOffice2013}. To further complication matters, if a competent proliferator uses shielding or naturally occurring radioactive material within the shipment, it is effectively impossible to pull out the SNM signal with the passive detection systems currently employed without generating significant false positives \cite{UnitedStatesGovernmentAccountabilityOffice2013,UnitedStatesCongress.House.CommitteeonHomelandSecurity.SubcommitteeonEmergingThreatsCybersecurity2007}.  Nonetheless, DNDO continues to invest significant effort into a technical solution that is at best 90\% effective, and at worst almost useless, with the goal of deterring an extremely low probability event.  The proposed policies in this report follow that same trend and fail to learn the lessons of history in this regard.

@@

\textbf{Alpha Radiation:} The radioactive decay of a nuclide which results in the emission of a 4-Helium atom. Alpha radiation is sometimes followed by gamma emission.

@@

\textbf{Special Nuclear Material:} Plutonium, 233-Uranium, Uranium with enrichments of 233-Uranium or 235-Uranium.

>>>>

\usepackage[ampersand]{easylist}

@@

\subsection{IAEA Monitoring}

@@

\subsection{UNSCR-1540 Nuclear Security Standards}

Existing Policy:  \textcolor{red}{UNSCR-1540 Nuclear Security Standards}

Proposed Policy:  \textcolor{red}{Export nuclear security training}

Lead Agency: DOT, DoD

Contributing Agency: NNSA  \normalfont

The proliferation of  WMD is a serious national and international security threat. A possible pathway of ISIL obtaining SNM and complete warheads is theft.  UNSCR-1540 is directed at maintaining peace by preventing the proliferation of  WMD. This resolution proposes key actions aimed at non-proliferation \cite{Gomes2007a}:

\begin{easylist}[enumerate]
& Withholding any supportive actions of non-UN-member States having chemical, biological and nuclear weapons.
& UN members should have appropriate laws that prevent  non-member States from development, manufacturing, obtaining, transportation, and usage of nuclear weapons.
& States should have an appropriate system of internal control for non-proliferation of chemical, biological, and nuclear weapons, including control over the crucial materials for the WMD. In order to achieve these goals, States should develop:
&& A system to effectively account for items and materials related to  chemical, biological, and nuclear weapon processing, usage, transportation, and storage.
&& A system of physical protection for items and materials related to chemical, biological, and nuclear processing, usage, transportation, and storage.
&& A control system to detect, stop, and combat (within international law if needed) the illegal transportation of  items and/or materials for chemical, biological, and nuclear weapons.
& Creation of a Committee of the Security Council that will inspect the execution of the current resolution and will report the results to the Security Council.
& Recommends all UN member States to:
&& Put more effort on improvement of multilateral treaties about prevention of  chemical, biological, and nuclear weapons.
&& Accept national regulations for prevention of proliferation, which will have compliance with the multilateral  and international non-proliferation treaties.
& Recommends the States to be more active in nonproliferation cooperation aimed to prevent illegal trafficking of chemical, biological, and nuclear weapons, related items and materials, and their methods of delivery.
& Recognizes that some States might want to request  assistance for execution of the above resolutions. Assistance might be in terms of skills and resources that are  crucial to achieve the provisions above.
\end{easylist}

% Proposed Policy: Export Nuclear Security Training

Implementation of these rules would limit the proliferation of WMDs. However, there might be States with nuclear weapons where nuclear facilities have low physical security levels. These facilities are vulnerable to the theft of nuclear weapons and/or materials by ISIL, an organization that has the military and financial capability to successfully complete such an operation.

Global Security estimates the total assets of ISIL at \$500 billion \cite{GlobalSecurity}. Their  military capacity is estimated to include about 31,000 fighters \cite{BBC2014}. To prevent risk of proliferation, actions toward security improvements should be taken. First of all, highly vulnerable nuclear facilities that contain nuclear warheads, materials, and technology should be given the same defense as in Provision \#7 of the UNSCR-1540 \cite{Gomes2007a}.

Furthermore, the Committee of the Security Council should be given greater enforcement power to bring together the international community in order to respond quickly to the threat of ISIL. It should also engage in training  to equip nuclear states with low security and highly vulnerable nuclear facilities. Security training would include personnel training and the use of surveillance and other technologies to prevent  unauthorized access.

A barrier to assessing the necessary level of security training  is that IAEA does not know the condition of those facilities in countries that have not signed the NPT. Pakistan, India, and Israel do not allow IAEA inspectors to investigate or visit their nuclear facilities, thus the security training cannot be catered to the specific country.  Therefore, nuclear training can occur in two ways:

\begin{easylist}[enumerate]
& International Atomic Energy Agency (IAEA) runs over 60 nuclear security and education training events each year \cite{InternationalAtomicEnergyAgency2014}. These events cover the different aspects of nuclear facility security, including physical protection, identification of weak points of nuclear facility security, and information security.
& Exporting security training to the nuclear facilities of the countries that request assistance, according to Provision \#7 under the USCR-1540 \cite{Gomes2007a}. This will allow for more accuracy in giving proper training to the countries, by tailoring it to the special features of the locations.
\end{easylist}

We suggest that the United Nations  extend UNSCR-1540 to adding provisions about nuclear security training. We must also provide incentives to UN members and non-UN-member countries to attend the nuclear security educational events hosted by IAEA. Moreover, training needs to heavily focus on managing and securing  chemical, biological, and/or nuclear weapons materials and facilities.

% Budget and Implementation Time

The IAEA's Education and Training in Nuclear Security is a subgroup of IAEA's Nuclear Safety and Security. The budget of  Nuclear Safety and Security is about 37 million Euros  as of 2015 \cite{Iaea2014}. Since these trainings are held annually by the IAEA, there is no need for additional financing for visitor countries. However, for export training, additional financing is required. Even so, financing should not exceed more than a few hundred thousand dollars per year, per training.

The time for the implementation of an  expert group specializing in nuclear training can range from six months to a year, based on past experience.  Training groups can be built from the same groups in the IAEA that hold nuclear security and education events. The U.S Department of Transportation and NNSA (National Nuclear Security Administration) will work in conjunction with countries to help them implement safety across borders to prevent the movement (theft) of nuclear material.

\subsection{Global Threat Reduction Initiative} \label{sec:GTRI}

Existing Policy:  \textcolor{red}{Global Threat Reduction Initiative}

Proposed Policy:  \textcolor{red}{Renew and broaden GTRI commitment}

Lead Agency: DOS

Contributing Agency: DTRA, DoD, NNSA, NPWG  \normalfont

GTRI existing policy :

GTRI prevents nuclear material from getting into the wrong hands by: removing material from improperly protected places, converting nuclear  reactors from highly enriched uranium (HEU) to low enriched uranium (LEU), and improving security at nuclear facilities that are vulnerable to theft or sabotage [1]. Thus far, 49 HEU research reactors have been converted or shut down. Currently, security enhancement involves installation of detectors, such as Passive Infrared Sensors (PIR), Remote Monitoring System (RMS), Personal Radiation Detector (PRD), etc [2]. Courses have been developed to train personnel of nuclear facilities on how to use the advanced detectors and equipment to secure their nuclear facilities, as well as how to respond in case of potential theft or sabotage. GTRI should provide travel funding for people to attend courses where they can get a theoretical basis and hands-on security experience. GTRI is also collaborating with universities to create a nuclear security curriculum.

Proposed policy

The GTRI should continue to provide assistance to domestic nuclear facilities to remove unused nuclear material, enhance security in storage and transport of nuclear material, and increase training to local law enforcement.

The GTRI should put emphasis on establishing collaboration with international organizations and countries throughout the world to enhance nuclear security.  In addition to working directly in foreign countries, GTRI should provide training and equipment to local state agencies to enable them to assure the safety of their nuclear facilities.

Russia still houses the most radioactive material and has dozens of nuclear reactors that still use HEU fuel.  Bilateral agreement with countries like Russia should be established in addition to the GTRI to allow close adjustment to the specific situation.

Budget and Implementation

Converting or possibly shutting down the 200 HEU research reactors is expected to be completed by 2035, under the current budget plan.  By 2018, 98\% of excess and vulnerable weapon grade material will be removed[3]. 8,500 buildings with will be secured with high-priority nuclear and radiological material is predicted to be finished in 2044[4]. The budget for fiscal years 2012, 2013, and 2014 are \$503453, \$501048 and \$424487.

% Sources
% [1] GTRI: Reducing Nuclear Threats, National nuclear security administration, http://nnsa.energy.gov/mediaroom/factsheets/reducingthreats
% [2] Global Threat Reduction Initiative, nuclear and radiological material protection,  NNSA fact sheet, http://osrp.lanl.gov/Documents/NNSAGTRIfact%20sheet.pdf
% [3] Department of Energy FY 2014 Congressional Budget Request, DOE/CF-0084, Volume 1, National Nuclear Security Administration Office of the Administrator Weapons Activities Weapons Activities Defense Nuclear Nonproliferation Naval Reactors, April 2013
% [4] US Budget Request Shows Eroding Focus on Nuclear Security, Nickolas Roth, March 25, 201l

\subsection{Bilateral Nuclear Security Agreements}

Existing Policy:  \textcolor{red}{Bilateral Nuclear Security Agreements}

Proposed Policy:  \textcolor{red}{????????}

Lead Agency: DOS

Contributing Agency: ??????????  \normalfont

Russia

The U.S. signed the New Strategic Arms Reduction Treaty (New START) with Russia on 8 April 2010, and the treaty entered into force on 5 February 2011. The treaty limits the number of deployed strategic (SNW) to 1,550 on each side, to be achieved by February 2018. Deployed ICBMs, SLBMs, and heavy bombers assigned to nuclear missions are limited to 700, and deployed and non-deployed ICBM launchers, SLBM launchers, and bombers are limited to 800. Non-deployed ICBMs and SLBMs will also be monitored through on-site inspections and must be located at specified facilities away from deployment sites; labeled with unique identifiers [1].

New START provided a verification mechanism that allow both sides to keep check on each other's nuclear capabilities, building transparency and confidence. Importantly, the treaty will also keep account of all SNM/warheads that exists, not only for the benefit of US-Russia nuclear relations but also the possibility of nuclear weapons proliferation at the hands of non-state actors like ISIL.

Policy recommendations:

The US should maintain New START as one of the options to strengthen the overall verification process of nuclear weapons, missiles and launchers as means to ensure global nuclear weapons security. New Start has been under the purview of the US Department of State, working in conjunction with the US Department of Defense.

Up to 2012,  the cost of implementation of the New START has been \$301 million USD [2]. Although the New START is not put into effect with the aim of preventing ISIL, it will have have positive complementary effects. Presumably, it will make it more difficult for ISIL to obtain and transport nuclear weapons and material through theft because of increased nuclear security.

Pakistan

One of the most troublesome aspects of Pakistan's highly vulnerable nuclear security facilities is the high likelihood of theft by ISIL. Although the US has been very supportive of Pakistan's efforts to ensure its nuclear weapons security, there are various challenges that could still undermine Pakistan's efforts. Pakistan has received various military and non-military aid, close to 28.4 billion USD post 9/11 [3].

Policy recommendations:

The US need to continue to assist Pakistan in increasing its nuclear security capabilities, by providing monetary assistance and military hardware, as well as services aid like training and intelligence-sharing. The U.S. Department of State has been the lead agency in maintaining the US-Pakistan bilateral relations; however, when military specific assistance is needed, DoS will work closely with the U.S. Department of Defense on this issue.

North Korea

North Korea's nuclear capabilities and hostile leadership are distressing elements with respect to the growing threat of ISIL and global efforts to achieve nuclear weapons security. Because of that, North Korea's lack of cooperation poses one of the largest threats to ensuring that ISIL does not acquire, transport, or trade nuclear weapons.

The absence of diplomatic relations with North Korea stems from North Korea's attempted annexation of South Korea resulting in the Korean War and US imposed economic sanctions on North Korea [4]. The only communication between the two countries has been under the Six-Party Talks for the Denuclearization of the Korean Peninsula. Most forms of U.S. economic assistance, other than purely humanitarian assistance, are prohibited; therefore, U.S. assistance has been limited to food or energy aid. Between 1995 and 2001, the US has spent more than \$1.3 billion USD in total aid to North Korea (food aid - \$708 million USD , energy assistance - \$404 million USD, Six-Party Talks-related assistance - \$191 million USD) [5].

Policy recommendations:

The U.S. should continue the Six-Party Talks and begin actively engaging North Korea. Historical memory and isolation from the international world has made North Korea extremely hostile to the U.S. However, North Korea desires international legitimacy and has claimed that it wants to be a responsible nuclear power. Building upon that objective, North Korea and the U.S. can form multilateral agreements with China and South Korea aimed at preventing theft or sale of weapons through securing nuclear weapons facilities. Eventually, conditional agreements can be put in place where North Korea allows IAEA inspectors in exchange for food or energy.

Iran

The U.S. also need to address Iran's nuclear ambitions.  The U.S. is working under P5+1 initiatives to negotiate a nuclear deal to halt Iran's nuclear development. This cooperation will ensure monitoring and verification of Iran's nuclear development. Although it is known that Iran and ISIL have opposing ambitions and will likely not cooperate together, Iran's nuclear capabilities could also lead to proliferation of nuclear weapons into non-state actors and future proxy wars involving nuclear weapons.

Policy recommendations

The U.S. should work with Iran to prevent ISIL's acquisition of nuclear weapons. Easing economic sanctions and welcoming Iran into the international community will help facilitate bilateral relations for the mutual goal of defeating ISIL.

% References:
% [1] New Start, US Department of State (http://www.state.gov/t/avc/newstart/)
% [2] Report on Plan to Implement the Nuclear Force Reductions, Limitations, and Verification and Transparency Measures Contained in the New START Treaty Specified in Section 1042 of the National Defense Authorization Act for Fiscal Year 2012 (U)
% [3] Upadhyay, Brajesh, ‘US and Pakistan: Will the relationship endure?', BBC News, 19 Nov 2014 (http://www.bbc.com/news/world-us-canada-30105416)
% [4] US Relations with North Korea Fact Sheet (http://www.state.gov/r/pa/ei/bgn/2792.htm)
% [5]Manyin, Mark E. and Nikitin, Mary Beth, ‘Foreign Assistance to North Korea', Congressional Research Service, 2 April 2012
% [6] The Enemy of Our Enemy: Iran and ISIS on a Collision Course (http://www.breitbart.com/national-security/2014/09/17/the-enemy-of-our-enemy-iran-and-isis-on-a-collision-course/)

@@

RPMs evolved out of a layered defense concept to prevent a low-probability/high-impact scenario, nuclear terrorism. This layered defense is composed of RPMs as discussed in \autoref{sec:domestic_protection}, the overseas equivalent known as the SLD as discussed in \autoref{sec:Trafficking}, and the Global Threat Reduction Initiative discussed in \autoref{sec:GTRI}.  Despite the layered technical and diplomatic approach, there are significant gaps in the ability to intercept a shipment into the U.S. containing SNM.  For example, DHS has implemented 99\% screening at seaports, but struggles to implement a significant screening at airports, due to a lack of natural choke points and disruption caused by screenings, and at the U.S.'s notoriously porous border crossings \cite{DepartmentofHomelandSecurityDHS2013,GovernmentAccountabilityOffice2012}.  Even if a shipment does cross a border, seaport, or airport with a RPM, the detection probability is only 90\% in the best of scenarios, if the RPM is even operational at the time \cite{DepartmentofHomelandSecurityDHS2013,DepartmentofHomelandSecurityDHS2006}.  To top of the list of issues, the ANSI standard for certification of RPMs contains no SNM requirements due to the inherent difficult in passive detection of SNM \cite{Ansi2007}.  The ASP program was conceived to address some of these problems and increase interception through advanced detectors and algorithms, but it failed to pass 6 of 10 criteria in field tests, criteria which were widely criticized as being favorable to the vendors in the first place \cite{UnitedStatesGovernmentAccountabilityOffice2013}. To further complication matters, if a competent proliferator uses shielding or naturally occurring radioactive material within the shipment, it is effectively impossible to pull out the SNM signal with the passive detection systems currently employed without generating significant false positives \cite{UnitedStatesGovernmentAccountabilityOffice2013,UnitedStatesCongress.House.CommitteeonHomelandSecurity.SubcommitteeonEmergingThreatsCybersecurity2007}.  Nonetheless, DNDO continues to invest significant effort into a technical solution that is at best 90\% effective, and at worst almost useless, with the goal of deterring an extremely low probability event.  The proposed policies in this report follow that same trend and fail to learn the lessons of history in this regard.

@@

\textbf{Alpha Radiation:} The radioactive decay of a nuclide which results in the emission of a \ce{^{4}He} atom. Alpha radiation is sometimes followed by gamma emission.

@@

\textbf{Special Nuclear Material:} Plutonium, \ce{^{233}U},  or \ce{^{235}U}.
